Regional Base-Flow Index in Arid Landscapes Using Machine Learning and Instrumented Records

### 2025-05-28 elsevier_MANUSCRIPT.qmd
updates to full text

  - name: Benjamin Lucas
    orcid: 0000-0002-2021-3076
    corresponding: false
    roles: 
      - Conceptualization
      - Resources
      - Supervision
      - Software
      - Writing – Review & Editing 
    affiliations:
      - 'Department of Mathematics and Statistics, Northern Arizona University, Flagstaff, AZ, USA'
abstract: |
  Base flow, sustained by groundwater discharge, is a vital component of river ecosystems, particularly in drylands, where water resources are limited. This novel study analyzes the instrumented streamflow record in Arizona to assess long-term base-flow index (BFI) trends across gauged catchments. Results indicate that approximately 32% of Arizona’s streamflow originates from groundwater discharge, with significant spatial variability driven by landscape and climatic factors. Base-flow relationships are analyzed with coincident trends in climate variables such as precipitation, evapotranspiration, and temperature. Spatial and climatic trends reveal  variability in base-flow contributions, providing insight into groundwater-surface water interactions in arid and semi-arid landscapes. Building on this analysis, we applied machine learning methods to predict BFI in ungauged basins, addressing the challenges of Arizona’s sparse streamgage network. Using the eXtreme Gradient Boosting (XGBoost) algorithm trained on hydroclimate and physiographic predictors, we estimated long-term BFI from 1991 to 2020. This combined approach integrates observational data with predictive modeling to enhance our understanding of base-flow processes and provide a framework for water resource management in data-limited regions.

bibliography: references.bib
---

@@

Despite advancements, existing large-scale studies have limited applicability to the unique hydrogeologic conditions in arid and semi-arid regions like Arizona, leaving uncertainty regarding groundwater-surface water interactions in these critically water-stressed areas. This study addresses this crucial gap by developing a regionally tailored machine learning model specifically designed to estimate BFI for Arizona's ungauged basins. Utilizing hydrogeologic characteristics and hydroclimatic data, we estimate long-term mean BFI (1991–2020) in ungauged catchments, filling spatial gaps in the sparse streamgage network. Furthermore, regional trends in base flow and BFI at instrumented sites are analyzed alongside coincident climate trends in precipitation, reference evapotranspiration (ETO), and temperature. The outcomes offer novel insights into groundwater contributions to streamflow, supporting water resource management in an area increasingly vulnerable to drought and climate change impacts, and contributing broadly to the understanding of hydrologic processes in dryland environments. The objectives of this study are to (1) quantify long-term spatial and temporal trends in BFI across Arizona, and (2) develop a predictive model using eXtreme Gradient Boosting (XGBoost) to estimate BFI in ungauged basins based on hydroclimatic and geospatial variables.

@@

The implementation of machine learning models to predict hydrologic indices has been successful in past studies [@singh2018; @schmidt_ML_2020; @Rozos2021Machine]. In this work, we used the eXtreme Gradient Boosting (XGBoost) algorithm [@chen2016xgboost] to predict BFI at ungauged locations using catchment characteristics as predictors [@tbl-predictors]. The XGBoost algorithm is a decision tree-based ensemble algorithm, which can be adapted for either regression or classification problems. This algorithm iteratively builds an ensemble of decision trees, where each tree corrects errors from previous trees to improve predictions [@chen2016xgboost]. Its efficiency, scalability, and robustness have made it increasingly popular in recent years, with successful applications in environmental modeling tasks such as streamflow forecasting [@szczepanek2022daily; @ni2020streamflow] and land use/land cover classification [@Georganos2018Very].

XGBoost operates by leveraging gradient boosting on decision tree algorithms, combining multiple low-variance models to produce a robust overall prediction. Gradient boosting works iteratively: the initial tree is trained on the target values, while subsequent trees are trained on the residual errors of the preceding tree. Each tree is assigned a weight based on its contribution to reducing error, and these weights are used to determine the influence of each tree in the final model. The ultimate prediction is made by aggregating the outputs of all $n$ weighted trees in the ensemble. In this study, the trained XGBoost model is used to predict BFI in ungauged catchments based on geospatial and hydroclimate predictor variables (@tbl-predictors). Certain features were further subdivided according to their areal coverage within each basin (e.g. land cover was divided into 16 subdivisions). This approach allowed the model to capture finer-scale spatial variability and improve predictive accuracy.

|   | Variable | Source | Geoprocessing |
|----|----|----|----|
| Hydroclimate | Precipitation\* | PRISM | Basin average |
|  | Mean Temperature\* | PRISM | Basin average |
|  | Reference Evapotranspiration\* | TerraClimate | Basin average |
| Geospatial | Elevation\* | DEM | Basin average |
|  | Area | DEM | Basin average |
|  | Slope | DEM | Basin average |
|  | Aspect | DEM | Percent areal coverage |
|  | Land Cover\* | NLCD | Percent areal coverage |
|  | Hydrologic Soil Group\* | SSURGO | Percent areal coverage |
|  | Geology | USGS | Percent areal coverage |
|  | Karst | USGS | Percent areal coverage |

: Basin-characteristic variables used as initial features in XGBoost model. Starred features are maintained in the final, dimensionality-reduced model. {#tbl-predictors}

Our training dataset consisted of 7,724 site-year observations of annual BFI, where each observation represents a single water year at a given streamgage. These observations were paired with 45 predictor variables derived from basin characteristics and climate data. To optimize the model's performance, we first conducted an exhaustive grid search combined with 5-fold cross-validation to identify the optimal hyperparameter values. The hyperparameters evaluated included the learning rate ($\eta$), minimum split loss ($\gamma$), maximum tree depth, minimum child weight, and the number of trees. While not an exhaustive list of all possible XGBoost hyperparameters, this range of values provided sufficient variation to ensure the selection of a high-performing model. The optimal hyperparameters were determined to be: 700 trees, a learning rate ($\eta$) of 0.05, a minimum split loss ($\gamma$) of 0.075, a maximum tree depth of 7, and a minimum child weight of 5. Using these values, the XGBoost model was trained on the dataset with 10-fold cross-validation.

$K$-fold cross-validation provides an unbiased estimate of a model's accuracy on unseen data, while also insuring against overfitting or underfitting. In this approach, the data are randomly divided into $k$ folds of equal size. The model is trained on $k-1$ folds and tested on the remaining fold, referred to as the validation set. This process is repeated $k$ times, with each fold serving as the validation set exactly once. Each iteration trains an independent model with the same hyperparameters but using a different subset of training data. By averaging the model performance across all $k$ folds, we achieve a robust and reliable estimate of predictive accuracy. Root mean squared error (RMSE) was used as the performance metric for both model optimization and evaluation. RMSE provided a consistent and interpretable measure of the model’s accuracy throughout the training and validation process.

![Schematic of $k$-fold cross validation. The dataset is randomly divided into $k$ stratified folds. Each fold serves as the validation set once, while the remaining folds are combined to create a training set for model development. Performance metrics for the test set are calculated and recorded, and this process is repeated for all $k$ folds.](images/k-fold_fig.png){#fig-k-fold fig-align="center" width="100%"}

#### Feature Selection {#sec-feat-selection}

Machine learning models are prone to overfitting, especially when provided with a large set of predictive features [@Ying_2019]. Overfitting can degrade model performance on unseen data and increase the demand for computational resources and memory storage [@feat_selec2017]. Dimensionality reduction offers a robust solution to these challenges and generally falls into two broad categories: feature extraction and feature selection.

Feature extraction involves transforming the original dataset into a lower-dimensional feature space. However, this process generates new features that often lack the physical interpretability of the original variables. In contrast, feature selection identifies a subset of the original features, preserving their physical meaning while improving model readability and interpretability [@feat_selec2017]. In this study, supervised feature selection was employed to reduce the number of predictors, which enhanced learning performance, reduced computational costs, and mitigated overfitting.

To begin, an initial model was trained using the full feature set of 45 predictors (@tbl-predictors). A feature selection method based on feature importance was then applied to identify and remove less relevant and noisy features. Feature importance scores quantify the contribution of individual features—either positively or negatively—to the model's predictions [@ML_interpretability2019]. In this analysis, SHAP (SHapley Additive exPlanations) values were used to compute feature importance scores [@SHAP_values].

SHAP values is a method to explain the prediction of an individual instance by calculating the contribution of each feature to that prediction. The method is based on coalition game theory and is discussed further in @SHAP_values. Here, SHAP values are used for global interpretation of feature importance and feature effects on the model. Global feature importance is produced by the absolute Shapley values of each feature across the dataset, providing a list of features in order of most to least important. Feature effects provide an indication of the relationship between the value of a predicting feature and its impact on the prediction.

![SHAP value plot of features used in final model. The number to the right of the feature name is the mean absolute SHAP value. Land cover features are indicated by the percentage of cover by each land cover type and soil types are defined by hydrologic soil group.](images/shap_summary.png){#fig-shap_values}

The ten most important features [@fig-shap_values] were selected based on their SHAP values and used to train a subsequent model [@tbl-predictors]. This refined model demonstrated improved performance and reduced computational time compared to the initial model. The final model, trained on this optimized feature subset, was ultimately used for the analysis presented here.

@@

Statistical analyses were conducted on annual BFI and base-flow values from instrumented streamgages to identify temporal trends using the Mann-Kendall nonparametric trend test [@kendall1970; @mann1945nonparametric]. This test detects monotonic trends in datasets that are non-parametric and assumes the absence of autocorrelation among observations. This test is widely used in studies of this nature [@ficklin2016; @ayers2019; @woodhouse_udall22].

To check for autocorrelation, we applied the Durbin-Watson test [@durbin1950testing], which revealed significant autocorrelation at four streamgages on an annual basis. Of these, only one streamgage (09486500 - Santa Cruz River at Cortaro, AZ) showed a significant trend in BFI. This streamgage was excluded from the trend analysis, as autocorrelation could inflate the variance of the Mann-Kendall statistic, potentially leading to biased trend estimates [@MK_Hamed1998]. Trends with a $\alpha \le 0.05$ are considered significant.

@@

Previous studies have examined base-flow regionalization and synthesis across various spatial scales, some overlapping with our study area. @beck2013 explored global patterns of BFI using streamflow observations extrapolated through an ANN. The U.S. streamgages included in that analysis were drawn from the interim Model Parameter Estimation Experiment (MOPEX) dataset of 1862 USGS gauges [@beck2013; @mopex_2006]. Only 10 of these interim gauges overlap geographically with our study area. Similarly, @ayers2022 assessed climatic influences on monthly base flow across the continental United States (CONUS), utilizing a subset of GAGES-II streamflow gauges. Although exact gauge locations from that study are unavailable, published figures suggest minimal geographic overlap with our research domain. @santhi2008 employed interpolation methods for regional base-flow estimation across CONUS, but specific gauge locations or data are not available, limiting direct regional comparisons.

The existing streamgage networks Arizona are limited for base-flow analyses by their design and monitoring objectives. While the number of streamgages has increased to meet regulatory imperatives, such as those driven by the Clean Water Act, many gauges prioritize peak flow monitoring rather than low-flow conditions critical for base-flow studies [@flood-control2020]. Newer non-USGS gauges, such as those installed by flood control districts (e.g., the ALERT system), are generally tailored for flood detection rather than monitoring sustained base flow. Recent USGS installations focus on addressing in-stream flow rights and future monitoring improvements, though low-flow capabilities remain limited.

@@

>>>>

  - name: Benjamin Lucas
    orcid: 0000-0002-2021-3076
    corresponding: false
    roles: 
      - Conceptualization
      - Resources
      - Supervision
      - Software
      - Writing – Review & Editing 
    affiliations:
      - 'Department of Mathematics and Statistics, Northern Arizona University, Flagstaff, AZ, USA'
abstract: |
  Base flow, sustained by groundwater discharge, is a vital component of river ecosystems, particularly in drylands, where water resources are limited. This novel study analyzes the instrumented streamflow record in Arizona to assess long-term base-flow index (BFI) trends across gauged catchments. Results indicate that approximately 32% of Arizona’s streamflow originates from groundwater discharge, with significant spatial variability driven by landscape and climatic factors. Base-flow relationships are analyzed with coincident trends in climate variables such as precipitation, evapotranspiration, and temperature. Spatial and climatic trends reveal  variability in base-flow contributions, providing insight into groundwater-surface water interactions in arid and semi-arid landscapes. Building on this analysis, we applied machine learning methods to predict BFI in ungauged basins, addressing the challenges of Arizona’s sparse streamgage network. Using the eXtreme Gradient Boosting (XGBoost) algorithm trained and validated on observed hydroclimate and physiographic variables, we estimate long-term BFI from 1991 to 2020. This combined approach integrates observational data with predictive modeling to enhance our understanding of base-flow processes and provide a framework for water resource management in data-limited regions.

bibliography: references.bib
format: docx
---

@@

To address these limitations, this study focuses on Arizona’s arid and semi-arid basins, where groundwater–surface water interactions remain poorly characterized due to sparse monitoring and regionally distinct hydrologic behavior. We develop a region-specific machine learning model using eXtreme Gradient Boosting (XGBoost) to estimate long-term BFI (1991–2020) across ungauged basins, incorporating basin hydrogeology and hydroclimate predictors. In parallel, we assess trends in BFI at instrumented sites alongside coincident changes in precipitation, temperature, and reference evapotranspiration. Together, these analyses provide new insight into the spatial patterns and temporal dynamics of baseflow in a water-stressed region increasingly affected by climate change. The study objectives are to (1) quantify long-term spatial and temporal trends in BFI across Arizona and (2) develop a predictive model to estimate BFI in ungauged basins based on geospatial and climate variables.

@@

Machine learning (ML) approaches have demonstrated strong utility in hydrologic prediction tasks, including the estimation of streamflow and base-flow indices [@singh2018; @schmidt_ML_2020; @Rozos2021Machine]. In this study, we used the eXtreme Gradient Boosting (XGBoost) algorithm [@chen2016xgboost] to estimate annual BFI at ungauged locations across Arizona using catchment characteristics as predictors [@tbl-predictors]. XGBoost is a gradient boosting algorithm that constructs an ensemble of decision trees, where each tree iteratively corrects errors from previous trees to improve predictive accuracy [@chen2016xgboost]. The model was trained on annual BFI observations in order to estimate annual BFI in ungauged basins. These annual BFI estimates were averaged over time to represent long-term BFI, consistent with the time-integrated nature of our response metric.

|              | Variable                      | Source | Geoprocessing     |
|--------------|-------------------------------|--------|-------------------|
| Hydroclimate | Temperature                   | PRISM  | Basin Mean        |
|              | Mean Precip                   | PRISM  | Basin Mean        |
|              | Reference ET                  | MODIS  | Basin Mean        |
| Geospatial   | Elevation                     | DEM    | Basin Mean        |
|              | Soil Type A                   | SSURGO | \% Areal Coverage |
|              | Soil Type C                   | SSURGO | \% Areal Coverage |
|              | Land Cover - Open Water       | NLCD   | \% Areal Coverage |
|              | Land Cover - Low Develoment   | NLCD   | \% Areal Coverage |
|              | Land Cover - Evergreen Forest | NLCD   | \% Areal Coverage |
|              | Land Cover - Herbaceous       | NLCD   | \% Areal Coverage |

: Basin-characteristic variables used in reduced XGBoost model. A list of initial predictor variables used in the full model is provided in the Supplemental Information. {#tbl-predictors}

The training dataset consisted of 7,724 site-year observations, where each observation represents annual BFI for a single water year at a given USGS streamgage. These observations were paired with 45 predictor variables derived from basin characteristics and climate data. Predictor selection was based on prior hydrologic studies, data availability across the study area, and pairwise correlation screening to minimize multicollinearity. The predictors included basin-averaged climate metrics (e.g., precipitation, temperature), topographic features (e.g., elevation, slope), and land surface attributes (e.g., land cover classes, soil type, karst presence).

We tested multiple model configurations, including versions with the full predictor set across the state, separate models for physiographic regions with full predictor set, and reduced subsets across the state, to evaluate the impact of different predictor combinations on model performance and to ensure robustness of the final selected model. Dimensionality reduction was performed to improve interpretability and reduce computational burden. We applied SHAP (SHapley Additive exPlanations) values [@SHAP_values] to assess global feature importance and rank predictors. SHAP values quantify both the magnitude and direction of each feature’s contribution to predictions. The ten most important features—based on mean absolute SHAP value—were retained in a reduced model, which showed improved predictive accuracy and efficiency over the full model and regional models. Sensitivity analyses tested region-specific models and smaller predictor subsets; however, the state-wide model using SHAP-prioritized features consistently outperformed alternatives.

![SHAP value plot of features used in final model. The number to the right of the feature name is the mean absolute SHAP value. Land cover features are indicated by the percentage of cover by each land cover type and soil types are defined by hydrologic soil group.](images/shap_summary.png){#fig-shap_values alt="SHAP value plot of features used in final model. The number to the right of the feature name is the mean absolute SHAP value. Land cover features are indicated by the percentage of cover by each land cover type and soil types are defined by hydrologic soil group."}

The model dataset was evaluated using 10-fold cross-validation, ensuring robust performance estimates and guarding against overfitting [@fig-k-fold] . In each iteration (fold), the model was trained on 90% of the data and validated on the remaining 10%, with folds rotated until all data had been used for validation in separate models. Root mean squared error (RMSE) was used as the primary performance metric. Model performance on unseen data was strong, with an overall R² of 0.764. Additional model validation results are presented in section 3.2.1.

![Schematic of $k$-fold cross validation. The dataset is randomly divided into $k$ stratified folds. Each fold serves as the validation set once, while the remaining folds are combined to create a training set for model development. Performance metrics for the test set are calculated and recorded, and this process is repeated for all $k$ folds.](images/k-fold_fig.png){#fig-k-fold alt="Schematic of $k$-fold cross validation. The dataset is randomly divided into $k$ stratified folds. Each fold serves as the validation set once, while the remaining folds are combined to create a training set for model development. Performance metrics for the test set are calculated and recorded, and this process is repeated for all $k$ folds." fig-align="center" width="100%"}

Hyperparameter tuning was performed through an internal grid search. The search evaluated combinations of key parameters including the number of trees, learning rate ($\eta$), minimum split loss ($\gamma$), tree depth, and minimum child weight. A full discussion of these hyperparametes can be found in @chen2016xgboost. The final model used 700 trees, $\eta$ of 0.05, $\gamma$ of 0.075, a maximum depth of 7, and a minimum child weight of 5. All parameters were selected for their consistent reduction in RMSE across validation folds.

@@

Statistical analyses were conducted on annual BFI values from instrumented streamgages to identify site-specific temporal trends using the Mann–Kendall nonparametric trend test [@kendall1970; @mann1945nonparametric]. This test detects monotonic trends in non-normally distributed data and assumes the absence of autocorrelation. This test is widely used in hydrologic trend studies [@ficklin2016; @ayers2019; @woodhouse_udall22].

To account for potential autocorrelation in annual BFI time series at each instrumented site, we applied the Durbin–Watson test [@durbin1950testing]. Four streamgages exhibited significant autocorrelation; these sites were excluded from the trend analysis to avoid inflated variance in the Mann–Kendall statistic and potential bias in trend detection [@MK_Hamed1998]. As all autocorrelated series were removed, pre-whitening using the Hamed–Rao method was unnecessary [@MK_Hamed1998]. Trends with a $\alpha \le 0.05$ are considered significant.

@@

Several previous studies have examined regional and continental patterns of BFI, offering useful context for our findings despite limited geographic overlap. @beck2013 developed a global map of BFI using an artificial neural network trained on 1,862 U.S. streamgages from the MOPEX dataset [@mopex_2006]. Only ten of these gauges overlap with our Arizona study area, limiting direct comparison. Nonetheless, the spatial distribution of BFI in Arizona from @beck2013 shows general agreement with our regional patterns. The higher BFI values reported by @beck2013 may reflect the influence of their global-scale ANN model and associated generalizations. Similarly, @ayers2022 examined base-flow dynamics using GAGES-II sites across CONUS, but their study included few gauges within Arizona. While exact locations are not specified, their published figures suggests regional agreement with our estimated BFI distributions. @santhi2008 employed interpolation methods for regional base-flow estimation across CONUS, but specific gauge locations or data are not available limiting comparison.\`\`\`\`

In terms of long-term trends, our results align directionally with @ayers2022 , who found consistent monthly declines in base flow from 1989–2019 across the U.S. Southwest. Although our analysis focused on long-term trends in annual BFI, we observed a similar imbalance: 16% of sites showed significant decreasing trends compared to 8% with increasing trends. While we do not directly replicate their monthly analysis, the directional agreement supports the broader conclusion that groundwater contributions to streamflow are declining in many parts of the region.

The existing streamgage networks in Arizona are limited for base-flow analyses by their design and monitoring objectives. While the number of streamgages has increased to meet regulatory imperatives, such as those driven by the Clean Water Act, many gauges prioritize peak flow monitoring rather than low-flow conditions critical for base-flow studies [@flood-control2020]. Newer non-USGS gauges, such as those installed by flood control districts (e.g., the ALERT system), are generally tailored for flood detection rather than monitoring sustained base flow. Recent USGS installations focus on addressing in-stream flow rights and future monitoring improvements, though low-flow capabilities remain limited.

@@

The combined use of long-term baseflow trend analysis and XGBoost modeling offers a transferable framework for understanding groundwater–surface water interactions in data-scarce regions globally. This approach is particularly relevant in dryland and seasonally variable climates, where direct baseflow observations are limited and hydrologic responses are strongly shaped by physiography and land cover. Critically, arid regions around the world remain under-instrumented, with sparse monitoring networks and limited long-term streamflow records, despite facing acute water scarcity and growing pressure on groundwater systems [@taylor2013]. The success of such models depends on the availability and quality of climate, topographic, and soil datasets, as well as the representativeness of streamflow records used in training. Regions with differing hydrogeologic settings, such as glacial terrains, humid catchments with deep weathered profiles, or regions with inter-annual snowpack storage, may require alternative predictor variables or model structures. For instance, in humid regions with dense vegetation and shallow groundwater tables, evapotranspiration dynamics may play a more dominant role in baseflow variability, requiring region-specific calibration. Despite these caveats, the generalizable structure of this modeling framework, along with the interpretability offered by SHAP-based feature selection, makes it well-suited for adaptation to other geographic contexts, especially where water resources are vulnerable to climatic and land-use change
